Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-20 paper/main.tex
move table from supplement to applications section
completely reformulate introduction

%###################################################################################################
\section{Introduction}
\acp{abm} have proven to be powerfull tools in the study of complex cellular
systems~\cite{Pleyer2023}.
They have been applied in various scenarios, such as cancer
research~\cite{Ghaffarizadeh2018,Cooper2020}, Immuno-Oncology~\cite{Karolak2021} or the modeling
of mechanical intercellular connections in plant cells~\cite{DeRybel2014}.
By treating cells as individual entities, these tools allow researchers to think in terms of
cellular properties and processes thus allowing for a better understanding of the biological
reality.

Despite their popularity, \acp{abm} face multiple challenges in practise.
The construction of a fully functional \ac{abm} simulation from the ground up constitutes an
intricate and laborious process, necessitating a comprehensive array of skills. It is therefore
common practise to reuse existing simulation tools, which, in turn, brings along a number of
other problems.
Even if the existing tool is designed to be generalizable, it often comes with a pre-defined set
of parameters that need to be specified.
Furthermore, as we have discussed in our previous work~\cite{Pleyer2023}, these tools were most
often not designed to be flexible but rather built for specific purposes thus limiting their
generalizability~\cite{Pleyer2025Flexibility}.

Typically researchers estimate the parameters of their model using methods such as Maximum Likelihood
or Maximum a Posteriori~\cite{Gbor2015,Banga2008,Ashyraliyev2009}, check for parameter
identifiability using methods like profile-likelihood or apply model reduction
techniques~\cite{Kreutz2013,Raue2014,Simpson2025}.
However, in the context of \acp{abm} this practice is almost never followed and instead only shown
that the generated results lead to the
desired phenomenon aligning with the observed biological reality.
This lack of systematic analysis of the involved parameters leaves a large gap of explainability of
the underlying models.
Consequently, it reduces the confidence in the constructed models and raises concerns with respect
to the validity of the obtained results.

In this work, we showcase how a systematic analysis of parameters can be performed using a
general-purpose mechanical model for rod-shaped bacteria.
Rod-shaped Bacteria present an interesting target system since their physical form provides more
surface information compared to other forms such as spherical cells.
Their mechanical properties have been studied in great detail
experimentally~\cite{Chatterjee1988,Takeuchi2005,IWAI2002}.
It was found that the MreB protein is the driving force in maintaining their distinct elongated
shape~\cite{Ursell2014}, which plays an important role in the dynamics of
growth~\cite{Billaudeau2017}, division~\cite{Harry2001} and their collective
behaviour~\cite{vanGestel2015,Balagam2015}.
Furthermore, it has been discovered that the rod can be bent thus introducing temporary elastic or
permanent plastic deformations~\cite{Amir2014_2}.
In addition, parameters such as growth rates have been found to vary within a given cell
population~\cite{Koutsoumanis2013} which highlights the need to incorporate this heterogeneity when
constructing models on the single-cell level.

Computational models for rod-shaped bacteria allow researchers to systematically analyze their
properties and infer emergent phenomena~\cite{Cho2007,Jnsson2005}.
Existing tools can describe growth, division and many effects such as colony formation and
expansion~\cite{Matyjaszkiewicz2017,Doumic2020} but they are limited in describing the
aforementioned plastic or elastic deformations of the rod. Furthermore, the literature on how to
estimate the parameters of such models directly from data is very sparse.

This work addresses some of these previously listed limitations.
We use microscopic image time series as inputs to obtain information about the spatial configuration
of the cells.
This information can be used as initial values for a numerical simulation and also serves as ground
truth.
By varying the parameters of the model and comparing the results to the microscopic data, we can
estimate and quantify the parameters of our model.
Our approach takes a general stance which incorporates effects that are observed within
\textit{B.Subtilis} and \textit{E.Coli}.
We deliberately distinguish between conceptually separate components within this framework to allow
for downstream reusability.
The developed computational methods can be combined to systematically estimate the parameters of our
model and are readily available as a Python package.

@@

@@

\begin{table}[H]
    \centering
    \begin{tabularx}{\textwidth}{l c c c c c c}
        \toprule
        % \multicolumn{6}{Growth Rates}\\
        Method & Cell $0$ & Cell $1$ & Cell $2$ & Cell $3$ & Cell $4$ & Cell $5$\\
        \midrule
        \ac{abm} & 0.0011527990 & 0.0014106040 & 0.0018761827 & 0.0016834959 & 0.0036106023 & 0.0015209642\\
        Direct   & 0.0014057073 & 0.0010807540 & 0.0023177727 & 0.0015120675 & 0.0020488179 & 0.0016726263\\
        % 0.0014057073 & 0.0010807540 & 0.0023177727 & 0.0015120675 & 0.0020488179 & 0.0016726263\\
        % direct & 0.0009785162 & 0.0012641432 & 0.0021701728 & 0.0015695889 & 0.0036255661 & 0.0018419186\\
        \bottomrule
    \end{tabularx}
    \caption{TODO}
\end{table}

>>>>

%###################################################################################################
\section{Introduction}

% \textbf{Restructure Introduction with these bullet points}
% \begin{itemize}
%     \item Problem 1: Parameter estimation of \acp{abm}: how to do at all?
%     \item Problem 2: Many \acp{abm} have very many parameters
%     \item Problem 3: No \ac{abm} framework with support for flexible rod-shaped bacteria
%     \item Literature Gap 1: noone really does parameter estimation for \acp{abm}
%     \item Literature Gap 2: also no methods do this
%     \item Literture  Gap 3: Some literature on flexible rods and some models but often not with
%         growth, reactions etc.
%     \item Why matter 1: No explainability without values of parameters.
%     \item Why matter 2: How can we be sure to actually model biological reality?
%     \item Why matter 3: Bacteria especially elongated are fundamental. Understanding them is
%         important.
%     \item What we do 1: Construct new mechanical model for rod-shaped bacteria
%     \item What we do 2: Develop methods to compare numerical results with data (for previous model)
%     \item What we do 3: Apply these methods in 3 scenarios covering different aspects
%     \item Why my contribution significant 1: Show that parameter estimation can be done
%     \item Why my contribution significant 2: Show how to treat heterogeneity and that we can map 1:1
%         of model to cells
% \end{itemize}

\acp{abm} are powerful toosl to study complex cellular systems~\cite{Pleyer2023}.
By treating cells as individual entities, these tools allow researchers to think in terms of
cellular properties and processes thus allowing for a better understanding of the biological
reality.
Their applications range from cancer research~\cite{Ghaffarizadeh2018,Cooper2020} over
Immuno-Oncology~\cite{Karolak2021} to the modeling of mechanical intercellular connections in plant
cells~\cite{DeRybel2014} which demonstrates their ability to capture complex emergent multicellular
behavior.
However, despite their popularity \acp{abm} face multiple practical challenges.

One major problem is how to effectively perform parameter estimations for \acp{abm}.
Classical approaches in systems biology—such as maximum
likelihood~\cite{Kreutz2013,Raue2014} or maximum a posteriori estimation, identifiability analysis
and model reduction~\cite{Simpson2025,Gbor2015,Banga2008,Ashyraliyev2009}.
While some attempts have been made to properly calibrate the \ac{abm} at hand and estimate its
parameters~\cite{An2016,Lima2021,Dancik2010,Thiele2014} these methods are not widely applied and are
generally not used to explicitly capture any possible heterogeneity within the cellular population.

Many \acp{abm} contain a substantial amount of parameters which poses another major problem
related to the tuning of the model and can limit it's biology interpretability.
Therefore, many studies only show that a suitably chosen set of parameters produces the desired
biological behavior and asses how a subset of parameters can effect their results.
However, this can leave important questions unanswered:
Which combination of parameters drive the observed phenomena?
Can these parameters be identified given a set of datapoints?
Is there a more simple, i.e. reduced model that can describe the same phenomena?
If \acp{abm} are not able to answer these questions, their results will significantly lack in
explanatory power and ability to link to the biological problem at hand.
In addition, biolgical \ac{abm} are typically designed with specific systems in mind and come with a
predefined set of parameters which need to be specified and are not easily adaptable.
This circumstance leads to problems when aiming to find a minimal set of identifiable parameters
that can describe the desired phenomena.

Within our approach we will utilize a new computational model of flexible, elongated bacteria which
is a mechanical setup that most \acp{abm} are not capable of describing.
So far, only a few computational studies have considered flexible rods~\cite{Duman2018,Joshi2019}
but their model can not be transferred directly to more complex biolgical systems which might
involve intracellular reactions, extracellular processes, growth and division.
This poses another problem in the computational study of these systems.

The aforementioned problems present scientific significant gaps.
First, no model can be explained or interpreted without knowing the roles and values of its
parameters or how they are constrained.
Furthermore, models which are not properly calibrated can not guarantee to accurately represent the
underlying biolgical reality beyond numerical coincidence.
Finally, elongated bacteria are fundamental biolgical organisms which can be seen as a precursor to
more complex shapes such as spiral bacteria.
Obtaining a mechanistic understanding of their mechanical properties is essential to understand
emergent phenomana such as the formation of biofilms~\cite{Li2025} or bacterial
swarming~\cite{Starru2007}.

In this work, we address these challenges and aim to further close the previously identified gaps.
We construct a novel mechanical model for flexible, elongated bacteria that captures growth,
division, elastic deformations of the rod and mechanical interactions between cells.
We develop computational methods which utilize segmented microscopic images to quantitatively
compare numerical results with biolgical data on a single-cell level.
With these methods we can estimate the parameters of our model in three scenarios, each probing
different aspects of our model with increasing level of complexity.

Our approach demonstrates that parameter estimation of \acp{abm} is indeed feasible, even for
complex models and when incorporating cellular heterogeneity.
The developed methods fill a methodological gap and present a novel way of fitting individual-based
models to data which can be readily adapted to various other scenarios.
We present a novel mechanical model for flexible elongated bacteria which seamlessly integrates into
the \texttt{cellular\_raza} framework, enabling researchers to extend and adapt it in future
studies.
In total, these contributions push for a more rigorous, interpretable and biolgically grounded
development process for \acp{abm} which provide deeper mechanistic insights into the nature of
cellular behaviors.

@@

\begin{table}[H]
    \centering
    \begin{tabular}{l c c c c c c}
        \toprule
        % \multicolumn{6}{Growth Rates}\\
        Method & Cell $0$ & Cell $1$ & Cell $2$ & Cell $3$ & Cell $4$ & Cell $5$\\
        \midrule
        Fit   & 0.001405 & 0.001080 & 0.002317 & 0.001512 & 0.002048 & 0.001672\\
        \ac{abm} Direct & 0.001152 & 0.001410 & 0.001876 & 0.001683 & 0.003610 & 0.001520\\
        \ac{abm} Imaging\\
        % 0.0014057073 & 0.0010807540 & 0.0023177727 & 0.0015120675 & 0.0020488179 & 0.0016726263\\
        % direct & 0.0009785162 & 0.0012641432 & 0.0021701728 & 0.0015695889 & 0.0036255661 & 0.0018419186\\
        \bottomrule
    \end{tabular}
    \caption{TODO Update and explain these values. And also add units!}
\end{table}

@@
